---
title: "Perform multi-site water balance simulations using LWFBrook90R"
author: "Paul Schmidt-Walter"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
With the new LWFBrook90R version 0.4.0, multisite-simulations are becoming more
flexible, through new possibilities for retrieving climatic data input from a
function, and to use the `output_fun` argument of `run_LWFB90()` with reference
to identifiers for soil, climate and parameters, when called from
`run_multisite_LWFB90()`. To demonstrate the potential of the new functionality,
multi run simulations will be performed that reproduce the data of the NFIWADS
database, containing monthly and yearly aggregates of water fluxes and soil
moisture state variables for the National Forest Inentory of Germany (NFI).

## Prerequisites and data

To reproduce the shown examples, please install LWFBrook90R version 0.4.0 or
higher. You can install the package directly from github.com using the package
'remotes':
```{r, eval = F}
remotes::install_github(repo="pschmidtwalter/LWFBrook90R",
                        ref = "v0.4.0",
                        build_vignettes=TRUE)
```

Now load the required packages:
```{r, message=FALSE}
library(LWFBrook90R)
library(data.table)
library(RSQLite)
```

The NFIWADS sample data is available as supplementary materials of
NFIWADS-publication at ZENODO. We download the .zip file to a temporary location
and extract the sqlite database file 'NFIWADS_input.sqlite', that contains the
input along with a sample of climatic data, to save it in the working directory:
```{r}
tdir <- tempdir()
tfile <- tempfile(tmpdir = tdir, fileext = ".zip")
download.file("https://zenodo.org/record/1491520/files/NFIWADS_supplementary_material.zip", tfile)
unzip(tfile,  exdir= tdir, overwrite=TRUE)
file.copy(file.path(tdir, "NFIWADS_input.sqlite"), getwd())
```

Before running the NFIWADS example, we will also run some small examples using
the sample data that comes with LWFBrook90R. Therefore, we load the climatic and
soil physical data, create standard `param_b90` and `options_b90` objects and
set up a `soil` data.frame, using a pedotransfer function:
```{r}
data("slb1_soil", "slb1_meteo")
opts <- set_optionsLWFB90(startdate = as.Date("2002-06-01"), enddate = as.Date("2002-06-02"))
parms <- set_paramLWFB90()
soil <- cbind(slb1_soil, hydpar_wessolek_tab(texture = slb1_soil$texture))
```

# Multi-site simulations: climatic data input from a function {#msite-clim}

A basic multi-site simulation using the function `run_multisite_LWFB90()` runs
through lists of `param_b90`, `climate`, and `soil`-objects, and evaluates the
specified parameter sets for each of the soil/climate combinations:
```{r}
test <- run_multisite_LWFB90(options_b90 = opts,
               param_b90 = list(beech = parms,spruce = parms),
               climate = slb1_meteo,
               soil = list(soil1 = soil, soil2 = soil, soil3 = soil))
str(test, max.level = 1)
```
The results are returned as a list of the single run objects, as returned by
`run_LWFB90()`. The list entries are named according to the names of the list
entries holding the individual `param_b90`, `climate`, and `soil` input objects.
The function can easily be set up to run the 10 samples from the NFIWADS input
database for which climatic data is available. However, simulating all sites
from NFIWADS database in this way is not possible, because such a large list of
`climate` data.frames does not fit in the memory. Fortunately, it is possible to
pass a function instead of a `data.frame` as `climate`-argument to
`run_LWFB90()`. Such a function can be used to create the `climate`-data.frame
from a file or database-connection within `run_LWFB90` or
`run_multisite_LWFB90()`. For `run_LWFB90`, we can provide arguments to the
function simply via the `...`-placeholder. For `run_multisite_LWFB90()`,
however, we need to pass arguments to the function (possibly with individual
values for individual site, e.g. a file name) via the `climate_args`-argument.

To demonstrate this mechanism, we set up the database connection to
'NFIWADS_input.sqlite', extract the site ids for which climatic data is
available and display the tables in the database:

```{r}
#set up connection
dbinput <- dbConnect(SQLite(), "NFIWADS_input.sqlite")

# list tables
dbListTables(dbinput) #list tables

#retrieve site_ids for which climatic data is available 
ids_clim <- dbGetQuery(dbinput, "select distinct id from climate_daily_int100_sample")$id
```

The climatic data resides in the table 'climate_daily_int100_sample'. All
variables are stored as integers and need to be divided by 100 for converting
them to the units required by `run_LWFB90()`. Accordingly, we define a function,
that retrieves the climatic data for a specific NFI site from a table in the
database, does the data processing and returns a suitable `climate` data.frame:

```{r}
climfun <- function(con, tbl, site_id){
  #get data from connection con
  qry <- paste0("select * from ", tbl, " where id = '", site_id, "';")
  clim <- data.table(dbGetQuery(con, qry))

  # process data to be usable as 'climate' in LWFBrook90R::run_LWFB90()
  # use suitable names
  setnames(clim, c("grhds", "rrds","sddm","tadm", "tadn", "tadx","wsdm"),
           c("globrad", "prec","vpd","tmean", "tmin", "tmax","windspeed"))
  # convert units
  clim[, c("globrad", "prec","vpd","tmean", "tmin", "tmax","windspeed") :=
         list(globrad/100, prec/100,vpd/100,tmean/100, tmin/100, tmax/100,windspeed/100)]
  
  # create date variable
  clim[,dates := as.Date(paste(year, month, day, sep="-"))]
  
  # calculate vappres from vpd
  clim[, es :=  ifelse(tmean >= 0,
                       6.1 * 10^( (7.5 * tmean) / (tmean + 237.2) ) ,
                       6.1 * 10^( (9.5*tmean) / (tmean + 265.5) ) )]
  clim[, vappres := ifelse( (es - vpd) <0, 0, (es - vpd)*0.1)] #vpd und es in hPa
  
  # return sorted
  clim[order(dates),list(dates, tmin, tmax, tmean, globrad, vappres, prec, windspeed)]
}
```

Lets test our function with the database connection set up earlier to see that
it works:

```{r}
head(climfun(con = dbinput, tbl = "climate_daily_int100_sample", site_id = ids_clim[1]))
```

Before we run our multi-site simulation, we need to set up lists of arguments
for our above defined function, one for each site, and store them together in a
list. We define a list for each site, holding the required arguments. Although
`site_id` is the only site-dependend parameter, we need to specify all
non-default arguments for each site here. We loop across our earlier retrieved
vector of ids and name the entries according to the site ids, so they will be
propagated to the output of the simulations.

```{r}
clim_args <- lapply(ids_clim, function(x) {
  list(con = dbinput, 
       tbl = "climate_daily_int100_sample", 
       site_id = x)
  })
names(clim_args) <- ids_clim
```

For the simulation, we need to define the model control options, have to set up
the list of soil `data.frames`, and create the two parameter sets that we want
to run:
```{r}
# create LWFB90 options
opts <- set_optionsLWFB90(startdate = as.Date("2000-01-01"),
                          enddate = as.Date("2010-12-31"),
                          root_method = "soilvar",
                          budburst_method = "Menzel",
                          leaffall_method = "vonWilpert")

# retrieve soil data as data.table for ids_clim from database 
soils <- data.table(
  dbGetQuery(dbinput, paste0("select * from soils where id in (", 
                             paste(paste0("'",ids_clim,"'"), collapse = ", "),
                             "); "))
)
# convert to data.table and split to list
soils_list <- split(soils, by = "id") 

# create NFIWADS parameter sets  
param_beech <- set_paramLWFB90(winlaifrac = 0.1,maxlai = 6,sai = 1,height = 30,
                               frintlai = 0.1,frintsai = 0.25,cintrl = 0.06,cintrs = 0.4,
                               fsintlai = 0.1,fsintsai = 0.6,cintss = 0.6,cintsl = 0.6,
                               alb = 0.18,albsn = 0.23,lwidth = 0.05,glmax = 0.0042,
                               budburst_species = "Fagus sylvatica",infexp = 0.66)

param_spruce <- set_paramLWFB90(winlaifrac = 0.8,maxlai = 5.5,sai = 1,height = 30,
                              frintlai = 0.12,frintsai = 0.14,cintrl = 0.2,cintrs = 0.4,
                              fsintlai = 0.12,fsintsai = 0.14,cintss = 0.6,cintsl = 0.6,
                              alb = 0.14,albsn = 0.14,lwidth = 0.004,glmax = 0.0035,
                              budburst_species = "Picea abies (frueh)",infexp = 0.66)
```

Now we can run the multi-site simulation across the ten sites, using the two
above defined parameter sets. The list of `soil` data.frames is created by
splitting the `soils` according to their site id:
```{r}
test2 <- run_multisite_LWFB90(options_b90 = opts,
               param_b90 = list(beech = param_beech,spruce = param_spruce),
               climate = climfun,
               climate_args = clim_args,
               output = -1,
               soil = soils_list,
               cores = 5)
str(test2, max.level = 1)
```

# Multi-Site simulations: redirect output to file or database

The example above showed how to avoid loading climatic data into the workspace
all at once, by evaluating a function on-the-fly to deliver and process the
data. However, the output of `run_multisite_LWFB90()` can also become very large
if many sites are simulated, and the returned data is selected to contain daily
values. To reduce the output of the single run simulations, it is possible to
provide functions for `run_LWFB90()` via its `output_fun`-argument, that perform
directly on the output of simulation. These can be used to do some custom
aggregations or write the results to a file, while suppressing the return of the
regular selected output (as specified by `output`-argument) using `rtrn_output =
F`. In `run_multisite_LWFB90()`, such output-functions and further arguments can
be easily passed via the `...` placeholder to `run_LWFB90()`. However,
`run_LWFB90()` until recently did not know anything about the calling
`run_multisite_LWFB90()` and the names of the parameter/soil/climate data sets
(as specified by the names of the input list entries) it was currently
processing. This was a problem, if the output function was defined to write its
results to a file, because no identifier for the climate/soil/parameters sets
was available inside `run_LWFB90()`. Since version 0.4.0, the names of the
current soil, climate, parameter objects are passed automatically from
`run_multisite_LWFB90()` to `run_LWFB90()` as objects `soil_nm`, `clim_nm`, and
`param_nm`. In this way, they are accessible to `output_fun`-functions within
`run_LWFB90()`.

To demonstrate this mechanism, create a folder 'output', and define a simple
function, that will writes the results of the `daily_output` objects of the
individual runs to files in that folder. The file names are concatenated from
the names of the climate, soil, and parameter-sets.

```{r}
dir.create("output")
outfun <- function(x, clim_nm, soil_nm, param_nm) {
  fname = paste0("output/output_", clim_nm,"_",soil_nm, "_", param_nm, ".csv")
  write.csv(x$daily_output, file = fname)
}
```

Now we can run the multi-site simulation as before, but suppress the regular
output of `run_LWFB90()` using `rtrn_output=FALSE`, and pass above defined
function:

```{r}
test3 <- run_multisite_LWFB90(options_b90 = opts,
               param_b90 = list(beech = param_beech,spruce = param_spruce),
               climate = climfun,
               climate_args = clim_args,
               soil = soils_list,
               rtrn_input = FALSE,
               rtrn_output = FALSE,
               output_fun = outfun,
               cores = 5)
```

Let's take a look to one of the single run results in the returned list:

```{r}
test3[1]
```
We see that the entries do not contain any simulation results, because we passed
`rtrn_output = FALSE` to `run_LWFB90()`. As well, the entry `output_fun` is
`NULL`, because `write.csv` has no return value when a file is specified. We can
see however that out attempt was successful, the simulation results were written
to our 'output'-directory:
```{r}
list.files("output")
```
Our output function did not use any external arguments, all information (data
and names of the current input objects to create a file name) was internally
generated within `run_multisite_LWFB90()`. Similarily, we could have written the
simulation results to a database. To do so, we just needed to set up a database
connection inside the function, and close it afterwards. Arguments for setting
up the connection can simply be passed to `output_fun` via the `...`
place-holder in the call of `run_multisite_LWFB90()`. We will learn how to do
this in the next section.

# Reproduction of NFIWADS database
 
Now that we showed that our simple `output_fun` function worked, we can create a
more complex function that calculates the aggregates of soil moisture status
variables and water fluxes for the NFIWADS database. The output created consists
of two tables containing monthly aggregates and aggregates for the vegetation
period. To reduce code of the function, we outsourced some of the code in
functions that can be found in the directory 'R-Functions'. 

We source the functions inside our function call (see below). The first function
(`aggr_swati`) aggregates the daily layer-wise soil moisture and fluxes
variables (`layer_output`-data.frame) to daily values, while the other functions
aggregate daily values to monthly or vegetation period representations, for
which the beginning and end day-of-year of the vegetation period have to be
specified as vectors covering the years of the simulation. The aggregation of
the water fluxes simply take the output object `daily_output`, while soil
moisture state aggregation functions are designed to use the returned daily
values from the first function `aggr_swati`. Referring to these functions, we
can define our `output_fun` to combine the results in two tables, and write them
to two different tables (>70 columns!) in a database:

```{r}
outfun_nfiwads <- function(x, clim_nm, soil_nm, param_nm, db_path) {

  # source helper functions
  source("R-Functions/aggr_swati.R", local = T)
  source("R-Functions/aggr_fluxes_daymon.R", local = T)
  source("R-Functions/aggr_states_soil_daymon.R", local = T)
  source("R-Functions/aggr_fluxes_vegper.R", local = T)
  source("R-Functions/aggr_states_soil_vegper.R", local = T)
    
  # gather variables from model-input for later use
  vpstart <- x$model_input$param_b90$budburstdoy
  vpend <- x$model_input$param_b90$leaffalldoy
  
  soil <- with(x$model_input$param, merge(soil_nodes, soil_materials, by = "mat"))
  
  swat.profile <- aggr_swati(x$layer_output, soil) # aggregate layer output
  
  out_monthly <- data.table::data.table(site_id = clim_nm, parset = param_nm,
                            fluxes_dailytomonthly(x$daily_output),
                            states_soil_daymon(swat.profile)[,-c(1,2), with = F])
  data.table::setnames(out_monthly, names(out_monthly), tolower(names(out_monthly)))
  
  out_vegper <- data.table::data.table( 
    site_id = clim_nm, parset = param_nm,
    fluxes_dailytovegper(x$daily_output,vp.start = vpstart, vp.end = vpend),
    states_soil_dayvp(swat.profile, vp.start = vpstart, vp.end = vpend)[,-c(1,2), with = F])
  data.table::setnames(out_vegper, names(out_vegper), tolower(names(out_vegper)))
  
  # write to db
  # open connection
  db_out <- DBI::dbConnect(RSQLite::SQLite(), db_path)
  if (DBI::dbExistsTable(db_out, "nfiwads_monthly")) {
    chk_m <-DBI::dbAppendTable(db_out, "nfiwads_monthly", out_monthly)
  } else {
    DBI::dbCreateTable(db_out, "nfiwads_monthly", out_monthly)
    chk_m <-DBI::dbAppendTable(db_out, "nfiwads_monthly", out_monthly)
  }

  if (DBI::dbExistsTable(db_out, "nfiwads_vegper")) {
    chk_vp <- DBI::dbAppendTable(db_out, "nfiwads_vegper", out_vegper)
  } else {
    DBI::dbCreateTable(db_out, "nfiwads_vegper", out_vegper)
    chk_vp <- DBI::dbAppendTable(db_out, "nfiwads_vegper", out_vegper)
  }
  #close
  DBI::dbDisconnect(db_out)
  
  if (chk_vp & chk_m) {
    return("Success!1!!")
  }
}
```

The database should be created first. Like the input
database, we use an SQLite file, which is created when setting up the
connection:

```{r}
db_out <- dbConnect(SQLite(), "output/nfiwads_results.sqlite")
```

We can test our function with one of the single runs from our `test2` multirun
simulation to see that it works:
```{r}
outfun_nfiwads(test2[[1]],
           clim_nm = "clim", soil_nm = "soil", param_nm = "test", 
           db_path = "output/nfiwads_results.sqlite")
```

The database now contains two tables, populated with data:

```{r}
dbListTables(db_out)
#remove test-tables
dbRemoveTable(db_out, "nfiwads_vegper")
dbRemoveTable(db_out, "nfiwads_monthly")
```

Now it is time to run the NFIWADS sample, after having cleared the previously
created test-tables. Again, we set `rtrn_output = F`, and also `rtrn_input = F`,
so that only some small piece of information about the runs are returned. Note
that even though the model input is not appended to the simulation results, it
is available to our `output_fun`-function (`x`-argument), where we extract
information on the beginning and end of the vegetation period and information on
soil data.

```{r, echo = F}
options(future.rng.onMisuse = 'ignore')
```

```{r}
test4 <- run_multisite_LWFB90(
  options_b90 = opts,
  param_b90 = list(beech = param_beech,spruce = param_spruce),
  climate = climfun,
  climate_args = clim_args,
  soil = soils_list,
  rtrn_input = F,
  rtrn_output = F,
  output_fun = outfun_nfiwads,
  db_path = "output/nfiwads_results.sqlite",
  cores = 5)
```
Die Simulation läuft sauber durch, jedoch gibt es auf meinem REchner ein
Warning, dass parallel erzeugte Zufallszahlen nicht unabhängig von einandern
sind. Wahrscheinlich hat das etwas mit den Datenbankverbindungen zu tun,
Zufalleszahlen werden in LWFBrook90R jedenfalls nicht erzeugt. Mit folgendem
Befehl lässt sich die Warnung abschalten:
```{r, eval = F}
options(future.rng.onMisuse = 'ignore')
```



